cardinalR: Generating interesting high-dimensional data structures

### 2025-04-20 paper-cardinalR.Rmd
Added intial introduction version

snedata, geozoo, mlbench, splatter

>>>>

The generation of synthetic datasets with well-defined structural properties is a cornerstone of evaluating and benchmarking algorithms across diverse fields, including machine learning, data mining, and computational biology. Researchers frequently require the ability to create data with specific dimensionalities, noise characteristics, and intricate underlying structures to rigorously test the performance and robustness of their methods. The R ecosystem offers a variety of packages for generating synthetic data, each with its strengths and focus. For instance, **`geozoo` (Schloerke, [Year of last commit on GitHub])** provides a rich collection of geometric objects, facilitating the creation and analysis of specific shapes, primarily in lower-dimensional spaces. **`snedata` (Melville, [Year of last commit on GitHub])** offers tools for generating simplified datasets tailored for evaluating dimensionality reduction techniques like t-SNE, often focusing on revealing low-dimensional embeddings of complex structures. Packages like **`mlbench` (Leisch & Dimitriadou, 2010)** provide a collection of well-known benchmark datasets, often with established classification or regression challenges. In the realm of single-cell omics, **`splatter` (Zappia et al., 2017)** excels at simulating complex biological data, capturing nuances like batch effects and differential expression.

However, a research gap exists in the comprehensive and flexible generation of high-dimensional data structures that seamlessly integrate fundamental geometric principles with sophisticated noise control and customizable cluster generation. While `geozoo` offers geometric primitives, it lacks direct mechanisms for extending these structures to arbitrary high dimensions with controlled noise or for generating clustered data based on these shapes with user-defined parameters like position, scale, and orientation. `snedata` focuses on creating abstract datasets for dimensionality reduction evaluation, not on building high-dimensional data directly from geometric layouts with flexible noise and clustering options. `mlbench` provides static benchmark datasets, limiting the ability to generate data with specific, user-defined geometric structures and noise profiles in high dimensions. Similarly, `splatter` is specialized for biological data simulation and does not offer the general framework for creating diverse geometric structures in arbitrary dimensions with fine-grained control over noise and cluster arrangements based on basic geometric shapes.

To address this gap, this paper introduces the `cardinalR` R package. This novel package provides a versatile suite of functions designed to generate customizable data structures in any number of dimensions, starting from fundamental geometric primitives. `cardinalR` offers key functionalities that go beyond the capabilities of existing tools by enabling users to: (i) construct high-dimensional datasets based on geometric shapes, with the ability to augment dimensionality through the addition of controlled noise dimensions; (ii) introduce tunable levels of background noise to these structures; and (iii) generate complex clustered data arrangements by leveraging basic geometric forms with precise control over their positions, scales, orientations, and sample sizes in arbitrary dimensional spaces. By providing these integrated functionalities, `cardinalR` aims to empower researchers to create more nuanced and challenging synthetic datasets tailored to the specific demands of evaluating algorithms in high-dimensional settings and complex spatial scenarios, bridging the gap between geometric foundations and the flexible generation of complex synthetic data.

This paper will detail the architecture and functionalities of the `cardinalR` package, demonstrating its utility through illustrative examples and highlighting its advantages over existing R packages like `geozoo`, `snedata`, `mlbench`, and `splatter` in the context of generating high-dimensional, geometrically-informed, and noise-controlled synthetic data.
